\documentclass[12pt, a4paper]{article}

% Preamble
\usepackage[utf8]{inputenc}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{listings}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{tocbibind}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows, positioning}
\usepackage{xcolor}

% Page Style
\geometry{a4paper, margin=1in}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{Project Report}
\fancyhead[R]{\today}
\fancyfoot[C]{\thepage\ of \pageref{LastPage}}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}

% Code Listing Style
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    frame=single,
    numbers=left,
    numberstyle=\tiny,
    backgroundcolor=\color{gray!10}
}

% Hyperlink Setup
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
    pdftitle={Project Report - AI Text Summarization},
    pdfpagemode=FullScreen,
}

% Title
\title{
    \vspace{2cm}
    \textbf{\LARGE AI-POWERED TEXT SUMMARIZATION SYSTEM}\\
    \vspace{0.5cm}
    \large SHAMBHURAJ GADHAVE
    \vspace{1.5cm}
}
\author{}
\date{\today}

\begin{document}

% --- Title Page ---
\begin{titlepage}
    \maketitle
    \thispagestyle{empty}
    \vfill
    \begin{center}
        \large A Comprehensive Project Report\\
        \vspace{0.5cm}
        Full-Stack Web Application for Intelligent Text Summarization
    \end{center}
\end{titlepage}

\newpage

% --- Table of Contents ---
\tableofcontents
\newpage

% --- Abstract ---
\section{Abstract}
This report provides a detailed overview of an AI-powered text summarization application designed to address the challenge of information overload in the digital age. The project is a full-stack web application that generates concise and coherent summaries from lengthy texts using artificial intelligence. The system architecture comprises a React-based frontend built with Vite, Tailwind CSS, and shadcn/ui components for a modern and responsive user experience, and a Node.js/Express backend for robust server-side logic and API management.

\vspace{0.5cm}

Key features include secure JWT-based user authentication, synchronous and asynchronous text summarization with AI integration, Redis-based caching for performance optimization, BullMQ job queue for handling long-running tasks, and comprehensive CRUD operations for managing summarization history. The backend leverages MongoDB for data persistence, Redis for caching frequently accessed summaries, and integrates with external AI services for text processing. The application is containerized using Docker and includes deployment configurations for cloud platforms.

\vspace{0.5cm}

This document covers the complete software development lifecycle, including problem analysis, system design, technology stack selection, database schema design, API endpoint documentation, implementation details, testing strategies, and potential future enhancements.

\newpage

% --- Problem Statement ---
\section{Problem Statement}
In the contemporary digital landscape, individuals across various domains—including academia, business, journalism, and research—face an unprecedented challenge: information overload. The exponential growth of digital content means that professionals and students must process vast amounts of text from articles, research papers, reports, news, and documentation on a daily basis.

\vspace{0.5cm}

\textbf{Key Challenges:}
\begin{itemize}
    \item \textbf{Time Constraints:} Manually reading and summarizing lengthy documents is extremely time-consuming, reducing overall productivity.
    \item \textbf{Information Retention:} The human brain has limited capacity to retain and process large volumes of information effectively.
    \item \textbf{Consistency:} Manual summarization can be subjective and inconsistent across different users or time periods.
    \item \textbf{Scalability:} Processing multiple documents simultaneously is impractical for individual users.
    \item \textbf{Accessibility:} Not everyone has the skills or time to create effective summaries that capture the essence of the original text.
\end{itemize}

\vspace{0.5cm}

There is a critical need for an intelligent, automated tool that can distill lengthy texts into their essential points while maintaining the core message and context of the original content. The solution must be fast, accurate, scalable, and user-friendly, with the ability to handle various text lengths and styles. Additionally, users require a system that allows them to manage their summarization history, track statistics, and access their summaries across different devices.

\newpage

% --- Objectives ---
\section{Objectives}
The primary objectives of this project are as follows:

\subsection{Functional Objectives}
\begin{itemize}
    \item To develop a full-stack web application for automated text summarization using AI technology.
    \item To implement a secure user authentication and authorization system using JWT tokens.
    \item To provide both synchronous (real-time) and asynchronous (background job) summarization options to handle texts of varying lengths.
    \item To integrate external AI services for generating high-quality, coherent summaries.
    \item To allow users to save, view, edit, and delete their summarization history.
    \item To implement comprehensive statistics tracking including word counts, compression ratios, and usage metrics.
    \item To provide a responsive and intuitive user interface that works seamlessly across devices.
\end{itemize}

\subsection{Technical Objectives}
\begin{itemize}
    \item To build a scalable backend architecture using Node.js and Express.js capable of handling concurrent requests.
    \item To implement Redis-based caching to improve performance and reduce redundant AI API calls.
    \item To utilize BullMQ for managing asynchronous job queues for long-running summarization tasks.
    \item To design a normalized MongoDB database schema for efficient data storage and retrieval.
    \item To implement proper error handling, validation, and security measures throughout the application.
    \item To containerize the application using Docker for easy deployment and scalability.
    \item To follow RESTful API design principles for clean and maintainable code.
\end{itemize}

\newpage

% --- Existing System ---
\section{Existing System}
Before the development of this application, users relied on several existing approaches for text summarization:

\subsection{Manual Summarization}
Users would manually read through entire documents and create summaries by extracting key points. This method, while accurate when done by skilled individuals, is extremely time-consuming and not scalable for processing multiple documents.

\subsection{Basic Online Tools}
Several web-based summarization tools exist that use simple extractive methods (selecting sentences from the original text). However, these tools often have limitations:
\begin{itemize}
    \item Limited text length support
    \item No user authentication or history tracking
    \item Poor quality summaries that lack coherence
    \item No customization options for summary style or length
    \item Lack of API integration for programmatic access
    \item No support for asynchronous processing of large texts
\end{itemize}

\subsection{Commercial AI Services}
Platform-specific AI services (like ChatGPT, Claude) offer summarization capabilities but require users to:
\begin{itemize}
    \item Switch between multiple applications
    \item Manually copy-paste content
    \item Pay for separate subscriptions
    \item Lack centralized history management
    \item Have no integration with existing workflows
\end{itemize}

\subsection{Limitations of Existing Systems}
\begin{itemize}
    \item No unified platform combining authentication, summarization, and history management
    \item Lack of performance optimization through caching
    \item No support for both real-time and background processing
    \item Limited customization and control over summary generation
    \item Poor user experience and interface design
\end{itemize}

\newpage

% --- Proposed System ---
\section{Proposed System}
The proposed AI-powered text summarization system addresses all limitations of existing solutions by providing a comprehensive, full-stack web application with the following features:

\subsection{Core Features}
\begin{enumerate}
    \item \textbf{User Authentication System:}
    \begin{itemize}
        \item Secure registration and login using JWT tokens
        \item Password hashing with bcrypt for security
        \item Session management and token-based authentication
        \item Protected routes for authenticated users only
    \end{itemize}
    
    \item \textbf{Dual Summarization Modes:}
    \begin{itemize}
        \item \textit{Synchronous Mode:} Real-time summarization for shorter texts (up to 5000 words)
        \item \textit{Asynchronous Mode:} Background job processing for longer texts using BullMQ
        \item Job status tracking with unique job IDs
        \item Queue statistics and monitoring
    \end{itemize}
    
    \item \textbf{Intelligent Caching:}
    \begin{itemize}
        \item Redis-based caching of generated summaries
        \item Cache key generation based on text content and parameters
        \item 24-hour cache expiration for optimal performance
        \item Significant reduction in AI API calls and costs
    \end{itemize}
    
    \item \textbf{Summary Management:}
    \begin{itemize}
        \item Create, Read, Update, Delete (CRUD) operations
        \item Personal summary history for each user
        \item Automatic word count calculation
        \item Compression ratio tracking
        \item Timestamp tracking for all summaries
    \end{itemize}
    
    \item \textbf{Statistics Dashboard:}
    \begin{itemize}
        \item Total summaries created
        \item Total words processed (original and summarized)
        \item Average compression ratio
        \item User-specific analytics
    \end{itemize}
    
    \item \textbf{Modern User Interface:}
    \begin{itemize}
        \item Responsive design using Tailwind CSS
        \item Beautiful UI components from shadcn/ui
        \item Smooth animations with Framer Motion
        \item Toast notifications for user feedback
        \item Intuitive navigation and user experience
    \end{itemize}
\end{enumerate}

\subsection{Advantages Over Existing Systems}
\begin{itemize}
    \item Unified platform for all summarization needs
    \item Personalized user experience with authentication
    \item Performance-optimized with caching and job queues
    \item Scalable architecture supporting concurrent users
    \item Modern, responsive UI/UX design
    \item Comprehensive history and statistics tracking
    \item Containerized deployment for easy scaling
    \item RESTful API design for potential third-party integrations
\end{itemize}

\newpage
